---
title: "R Notebook"
output: html_notebook
---


EDA

```{r}

accidents_ne_agg_eda <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),
pop = mean(POPESTIMATE),
age = mean(MEDIAN_AGE_TOT),
pov = mean(PopBelowPoverty),
inc = mean(Median_Household_Income_2019),
bus = mean(Number.of.business.establishments),
bridge = mean(Number.of.Bridges),
reswork = mean(Number.of.resident.workers/POPESTIMATE),
comtrans = mean(Percent.of.resident.workers.who.commute.by.transit),
wfoc = mean(Number.of.workers.from.other.counties.who.commute.to.work.in.the.county/POPESTIMATE),
reswh = mean(Number.of.resident.workers.who.work.at.home/POPESTIMATE),
reswcwc = mean(Number.of.resident.workers.who.commute.within.county/POPESTIMATE),
)

```
```{r}
accidents_ne_agg_pop <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),pop = mean(POPESTIMATE))

ggplot(data = accidents_ne_agg_pop) +
  aes(log(pop),log(n_acc)) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "County level population vs accidents",x="County Population",y="No.of accidents in 3 years") +
  geom_smooth(aes(color = State),se=F,method = "lm") 
```

```{r}
accidents_ne_agg_age <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),age = mean(MEDIAN_AGE_TOT))

ggplot(data = accidents_ne_agg_age) +
  aes(age,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "County level median age vs accidents",x="County Median age",y="No.of accidents in 3 years")   #geom_smooth(aes(color = State),se=F,method = "lm") 
```
```{r}

accidents_ne_agg_bus <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),bus = mean(Number.of.business.establishments))

ggplot(data = accidents_ne_agg_bus) +
  aes(log(bus),log(n_acc)) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "County level business establishments vs accidents",x="Number of business establishments",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 

```


```{r}

accidents_ne_agg_bridge <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),bridge = mean(Number.of.Bridges))

ggplot(data = accidents_ne_agg_bridge) +
  aes(bridge,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "Bridges vs accidents",x="Number of bridges",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 

```

```{r}
accidents_ne_agg_p_air <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),p_air = mean(Primary.and.Commercial.Airports))

ggplot(data = accidents_ne_agg_p_air) +
  aes(p_air,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "Primary airports vs accidents",x="Number of bridges",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 
```



```{r}

accidents_ne_agg_mf <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),mf = mean(m_f_ratio))

ggplot(data = accidents_ne_agg_mf) +
  aes(mf,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "County Level Sex ratio vs accidents",x="Sex Ratio",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 

```


#Average Daily frequency  by year and state

```{r}

accidents_ne_agg_pov <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),pov = mean(PopBelowPoverty))

ggplot(data = accidents_ne_agg_pov) +
  aes(pov,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "County Level Population below Poverty vs accidents",x="Population below poverty",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 

```

#Year- month wise accident frequency

```{r}

accidents_ne_agg_inc <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),inc = mean(Median_Household_Income_2019))

ggplot(data = accidents_ne_agg_inc) +
  aes(inc,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "County Level Median household income vs accidents",x="Median household income",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 

```

#Average daily 

```{r}

accidents_ne_agg_inc <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),inc = mean(Median_Household_Income_2019))

ggplot(data = accidents_ne_agg_inc) +
  aes(inc,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "County Level Median household income vs accidents",x="Median household income",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 

```

#Statewise Average daily variation in accident frequency over a month

```{r}

accidents_ne_agg_resw <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),reswork = mean(Number.of.resident.workers/POPESTIMATE))

ggplot(data = accidents_ne_agg_resw) +
  aes(reswork,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "County Level Resident workers vs accidents",x="Resident workers",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 

```

#2 Average daily accidents by day of the week

```{r}

accidents_ne_agg_reswcwc <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),reswcwc = mean(Number.of.resident.workers.who.commute.within.county/POPESTIMATE))

ggplot(data = accidents_ne_agg_reswcwc) +
  aes(reswcwc,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "Resident workers who commute within county vs accidents",x="Resident workers who commute within county",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 

```

#1 Number of accidents by hour of day (split by severity maybe?)

```{r}
accidents_ne_agg_reswh <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),reswh = mean(Number.of.resident.workers.who.work.at.home/POPESTIMATE))

ggplot(data = accidents_ne_agg_reswh) +
  aes(reswh,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "Resident workers who work at home vs accidents",x="Resident workers who work at home",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 
```



#2 Number of accidents by Weather Condition (split by severity maybe?)
```{r}
accidents_ne_agg_reswoc <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),reswoc = mean(Number.of.resident.workers.who.commute.to.work.in.other.counties/POPESTIMATE))

ggplot(data = accidents_ne_agg_reswoc) +
  aes(reswoc,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "Resident workers who commute to work in other counties vs no.of accidents",x="Resident workers who work at home",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 
```

#Distribution of accidents by temperature


```{r}
accidents_ne_agg_wfoc <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),wfoc = mean(Number.of.workers.from.other.counties.who.commute.to.work.in.the.county/POPESTIMATE))

ggplot(data = accidents_ne_agg_wfoc) +
  aes(wfoc,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "Workers from other counties who commute to work in the county vs accidents",x="Workers from other counties communting in the county",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 
```

```{r}

accidents_ne_agg_comtrans <- accidents_ne_agg_g %>% group_by(State,County) %>% summarise(n_acc=sum(n),comtrans = mean(Percent.of.resident.workers.who.commute.by.transit))

ggplot(data = accidents_ne_agg_comtrans) +
  aes(comtrans,n_acc) +
  geom_point(alpha = 0.3,aes(color = State)) +
  scale_fill_brewer(direction = -1) +
  labs(title = "Percent of resident workers who commute by transit vs accidents",x="Percent of resident workers who commute by transit",y="No.of accidents in 3 years") + geom_smooth(aes(color = State),se=F,method = "lm") 

```

